Predicting the effects of character displacement on food-web dynamics

### 2019-01-23 manuscript_ECD_model.Rmd
update repo with many changes

---
title: "Predicting the effects of character displacement on food-web dynamics"
author: "Matt Barbour"
output:
  pdf_document:
    df_print: paged
    fig_caption: true
    citation_package: natbib
    includes:  
      in_header: preamble-latex.tex
bibliography: references_ECD_supp_mat.bib
fontsize: 12pt
---

Keywords: competition; eco-evolutionary dynamics; ecological speciation; three-spine stickleback; zooplankton

## Abstract

Ecological character displacement is a key driver of diversification and adaptive radiation; however, there has been little work examining the consequences of character displacement on the dynamics of ecological communities. Given the growing recognition for the role of evolution in structuring ecological communities, understanding how major evolutionary processes impact ecological dynamics is increasingly important. Here, we build a mathematical model to predict how exploitative competition between two consumers for two resources affects the evolution of consumer attack rates, and, in turn, consumer-resource population dynamics. We then compared these predictions to results from a natural system, where a fish consumer (three-spine stickleback) has become locally adapted to different aquatic habitats with different resources (benthic vs. limnetic zones). Our model suggested that increasing ecological character displacement decreases the equilibrium density of resources, and with sufficient divergence, results in more oscillatory (destabilizing) consumer-resource dynamics. In support of this prediction, we found that the biomass density of zooplankton (resource) in lakes with two species of stickleback was only a fraction of that found in lakes with one species, varying from 4 to 41%. Moreover, we found that the biomass density of zooplankton was 2-fold more variable (CV) in lakes with two vs. one species of stickleback. These results suggest that ecological character displacement can eventually have a negative feedback on food-web persistence in the absence of other processes.

## Introduction

Ecological character displacement (ECD) is thought to be a key evolutionary process in the diversification of species and consequently the generation of biodiversity (Schluter 2000; Pfennig and Pfennig 2010; Stuart and Losos 2013). ECD describes the “process of phenotypic evolution in a species generated or maintained by [exploitative] resource competition with one or more coexisting species” (Schluter 2000). Over the past 40 years, a large body of theoretical (e.g. Lawlor and Smith 1976; Abrams 1986; Doebeli 1996; Taper and Chase 1985) and empirical (reviewed in: Schluter 2000; Dayan and Simberloff 2005; Stuart and Losos 2013) work has been generated to understand the scenarios under which exploitative competition for resources leads to the divergence of consumer foraging traits. Yet, although the process of ecological character displacement is driven by the interplay between competition and trait change, and hence is strongly dependent on resource availability, little of the work to date has explicitly considered how character displacement impacts other species within the community. Similarly, research on food-web dynamics has largely ignored the potential role of evolution in altering interactions and stability (Fussmann et al. 2007). Given the emerging evidence that evolutionary processes can drive ecological dynamics (Hairston et al. 2005, Post et al. 2008, Harmon et al. 2009, Schoener 2011, Rudman et al. 2015), understanding the ecological consequences of character displacement is a natural first step in investigating how evolutionary processes affect food-web dynamics.

Ecological character displacement has typically been considered to be a stabilizing factor in communities (Lawlor and Smith 1976). This is because divergence in foraging traits is thought to reduce the strength of interspecific competition between consumers. Reducing the strength of interspecific competition relative to intraspecific competition is a stabilizing force in consumer-resource systems. Consumer specialization on distinct resources suggests that there are niche differences between consumers. Mathematically, this is also often assumed to be the case, because using techniques such as evolutionary invasion analysis describe the ability for a consumer with a mutant allele to increase in abundance when rare, a demographic signature of niche differentiation. Therefore, if both consumers are able to invade when rare, then this suggests that they can coexist. On the other hand, relatively specialized consumers have also been shown to be a strongly stabilizing force in food webs (McCann et al. 2005). This is because these more specialized consumers may generate asynchrony in the dynamics of patchily distributed resources. This asynchrony in resources then permits consumers to move to habitat patches with higher resource densities after they have depleted the resources in a different patch. If, for example, the consumers were not specialized, then the resources would exhibit the same dynamics, so when the resources at low densities, consumers would be unable to switch to alternative resources, which would cause the consumer populations to crash, and oscillate more with their resources (McCann et al. 2005). Consequently, whether ecological character displacement will tend to stabilize (via asynchronous resource dynamics) or destabilize (via increased interaction strength) food webs is unclear.

On the other hand, ecological character displacement often results in the evolution of characters that enhance a consumer’s ability to capture resources, which tends to destabilize consumer-resource interactions (Murdoch et al. 2003; McCann 2011).

This occurs because increases in a consumer’s attack rate results in the resource being suppressed well below its carrying capacity, which if sufficient enough, can generate oscillations, and hence less stable, consumer-resource dynamics (Rosenzweig 1971; Murdoch et al. 2003). Experimental decreases in interaction strength between consumers and resources has been shown to promote coexistence between consumers and resources (Luckinbill 1973). Moreover, there is broad support that specialized consumers, since they are tightly coupled to their resources, tend to exhibit consumer-resource cycles, whereas this is not the case for generalists (Murdoch et al. 2002). Therefore, although evolutionary responses to exploitative competition may reduce the strength of interactions between consumers, they ultimately increase the strength of interactions with their resources which can destabilize the direct consumer-resource interaction (rather than the indirect consumer-consumer interaction). While prior models of character displacement have included more realistic functional responses and spatial relationships between consumers and resources, they have focused more on whether these factors alter patterns of character displacement rather than the effects of character displacement on the stability of the ecological community. As a consequence, it is currently unclear whether ecological character displacement has a net stabilizing or destabilizing effect on ecological communities.

Here, we address this knowledge gap by building a mathematical model to examine how ecological character displacement affects consumer-resource dynamics. We then compared the predictions from our model to field data from one of the most well-studied examples of ecological character displacement in a natural system, threespine stickleback in lakes of southern British Columbia. We specifically sought to address two questions: (1) How does exploitative competition between two consumers for two resources affect the evolution of consumer attack rates? (2) How does the evolution of consumer attack rates affect community stability?

@@

## Model of resource competition

I analyzed a continuous-time model of two consumers ($C_1,C_2$) competing for two shared resources ($R_1,R_2$):

$$\frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{21}(R_1)C_2$$
$$\frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-F_{12}(R_2)C_1-F_{22}(R_2)C_2$$
$$\frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{12}F_{12}(R_2)C_1-m_1C_1$$
$$\frac{dC_2}{dt}=e_{21}F_{21}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2$$

where $r_i$ represents the intrinsic growth rate of resource $i$, $K_i$ represents the carrying capacity of resource $i$, $e_{ji}$ represents the conversion efficiency of resource $i$ into consumer $j$, and $m_j$ represents the mortality rate of consumer $j$. $F_{ji}(R_i)$ represents consumer $j$'s feeding rate (i.e. functional response) on resource $i$. This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size, see @Taper1985 for an example). In the scenario where consumer feeding rates increase linearly with resource density, such that $F_{ji}(R_i)=a_{ji}R_i$ where $a_{ji}$ is the attack rate of consumer $j$ on resource $i$, then this model becomes MacArthur's model of resource competition (@MacArthur1972). The evolution of consumer attack rates in this model (and several extensions) have been analyzed in detail (@Lawlor1976; @Abrams1986). The general result of these analyses is that divergent character displacement in attack rates is an inevitable outcome of resource competition. <!-- In other words, the presence of a competitor causes consumers to evolve more specialized attack rates ($\frac{a_{ii}}{a_{ii}+a_{ij}}$) compared to when no competitor is present. -->

Here, I examine the effects of this inevitable divergence in attack rates on food-web dynamics. In particular, I examine how divergence in attack rates affect resource densities as well as the local stability of both resources and consumers. I do this by (i) deriving analytical expressions for the relationship between attack rates and food-web dynamics in different foraging scenarios and (ii) simulating the effects of competition on the eco-evolutionary dynamics of consumers and resources. For these simulations, I used an Adaptive Dynamics approach. At each evolutionary time step, I created a mutant consumer by randomly choosing a consumer and modifying its attack rate on one resource by either subtracting or adding a small constant <!-- (0.001 in the following simulations) -->with equal probability. The mutant's attack rate on the other resource was determined by the shape of the tradeoff (details below). Assuming this mutant consumer is rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource dynamics to reach a steady state. I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy (ESS, @Smith1973) or an evolutionary limit (e.g. $\frac{a_{ii}}{a_{ii}+a_{ij}}$ is constrained to a maximum of 1 and minimum of 0). I show that the consequences of character displacement on food-web dynamics are not intuitive and depend critically on the foraging scenario and tradeoff in consumer attack rates.

## Effect of character displacement on resource densities

I start by examing the effects of character displacement on the density of resources at equilibrium, which will also set the stage for a later analysis of food-web stability. In particular, I examine two different foraging scenarios: one where consumers forage for both resources simultaneously (e.g. @MacArthur1972) and one where they cannot because resources occur in distinct habitats (e.g. @Lawlor1976; @McCann2005). I say consumers have undergone divergent character displacement if there evolved attack rates are more specialized, defined as $|\frac{a_{ii}}{a_{ii}+a_{ij}}-0.5|$, when evolving with a competing consumer.

### Consumers forage for both resources simultaneously

An inherent assumption in MacArthur's model of resource competition is that consumers can forage for both resources simultaneously. In essence, this implies that both resources occur within the same habitat. To gain analytical insight to this scenario, I assume that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $m=m_1=m_2$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). While this scenario is arguably simplistic, it isolates the effect of character displacement rather than any dependence on inherent differences in consumers or resources. I show that when both consumers and resources are present that the density of resources at equilibrium are equivalent ($R=R_1=R_2$) and are determined by the following equation (*Mathematica* file with derivation available upon request):

$$\hat{R}=\frac{1}{a_{ii}+a_{ij}}\cdot\frac{m}{e}$$
Note that a key determinant of resource density is the consumer's total attack rate, $a_{ii}+a_{ij}$. Since the total attack rate is in the denominator, this implies that resource densities will decrease if evolution favors an increase in total attack rate. I assume that consumers are not 'Darwinian demons' (@Law1979) and thus the evolution of their attack rates are subject to certain constraints. I modelled these constraints such that $\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff (@Sargent2006). This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher, on average, than the extremes (when $n>1$, green line in Fig. \ref{fig:Tradeoff}) or, conversely, where the two extremes are higher, on average, than intermediate investments (when $n<1$, orange line Fig. \ref{fig:Tradeoff}). When $n=1$, the tradeoff function is linear, and all combinations of $a_{ii}$ and $a_{ij}$ have the same total attack rate (blue line in Fig. \ref{fig:Tradeoff}).

```{r Tradeoff, echo=FALSE, fig.cap="\\label{fig:Tradeoff}Tradeoff forms for consumer attack rates."}
tradeoff_plot <- bind_rows(concaveDN_df, linear_df, concaveUP_df) %>%
  ggplot(., aes(x=aii, y=aij, color = factor(n))) + 
  geom_line() + 
  xlab(expression(a[ii])) +
  ylab(expression(a[ij])) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)

@@

Previous work has shown that consumers undergo divergent character displacement regardless of the shape of the tradeoff (@Lawlor1976; @Abrams1986). Interestingly, I find that the shape of the tradeoff qualitatively affects the relationship between character displacement and resource density (Fig. \ref{fig:MacArthur_Resources}). For example, if consumer's are constrained by a linear tradeoff (blue lines), then there is no net change in total attack rate (Fig. \ref{fig:MacArthur_Resources}A) and character displacement has no effect on resource densities (Fig. \ref{fig:MacArthur_Resources}B). If the tradeoff is concave down (green lines), then resource abundances can actually increase under character displacement (Fig. \ref{fig:MacArthur_Resources}B). This is because the total attack rate of consumers is maximized at intermediate values ($a_{ii}=a_{ij}$) and decreases as consumers diverge (Fig. \ref{fig:MacArthur_Resources}A). Note that detecting this effect in nature would likely be difficult, since a concave-down tradeoff results in relatively small displacement relative to other tradeoff shapes (@Lawlor1976; green points in Fig. \ref{fig:MacArthur_Resources}A,B). When the tradeoff is concave up (orange lines), character displacement suppresses resource densities due to the increase in total attack rates (Fig. \ref{fig:MacArthur_Resources}A,B). Although the equation I derived for resource density was for the scenario where both consumers and both resources are present, it predicts well the density of resources when a single consumer reaches its ESS (triangles on respective colored lines in Fig. \ref{fig:MacArthur_Resources}B). This is because a single consumer evolves to be a generalist that has equal attack rates on each resource (triangles at 0.5 along x-axis in Fig. \ref{fig:MacArthur_Resources}A), resulting in equivalent resource densities.

>>>>

---
title: Predicting the effects of character displacement on food-web dynamics
author: 
- Matthew A. Barbour$^{1,2,\ast}$
# potentially use for 'date' for affiliations, as recommended by https://github.com/rstudio/bookdown/issues/348 for cross-format compatibility, with | to create new sections.
#date: | 
#    | 1. Affiliation
output:
  pdf_document:
    keep_tex: true
    toc: false
    number_sections: false
    df_print: paged
    fig_caption: true
    #citation_package: natbib
    template: Rmd_AmNat_MS_template.tex
    includes:  
      in_header: Am-Nat-preamble-latex.tex
bibliography: references_ECD_supp_mat.bib
biblio-style: amnat.bst
fontsize: 11pt
fontfamily: mathpazo
documentclass: article
linkcolor: black
urlcolor: black
---

<!-- Begin title page formatting -->

\noindent 1. University of British Columbia, Department of Zoology, Vancouver, BC V6T 1Z4, Canada;

\noindent 2. University of Zurich, Department of Evolutionary Biology and Environmental Studies, Winterthurerstrasse 190, 8057 Zurich, Switzerland;

$^\ast$ Corresponding author; e-mail: matthew.barbour@ieu.uzh.ch

\bigskip

*Manuscript elements*: Figure 1, figure 2, figure 3, figure 4. All figures should be printed in color.

\bigskip

*Keywords*: competition; eco-evolutionary dynamics; consumer-resource interactions; adaptive radiation; ecological community

\bigskip

*Manuscript type*: Note.

\bigskip

\footnotesize Prepared using an *Am. Nat.* inspired \LaTeX{} template for Rmarkdown. \normalsize

<!-- End title page formatting -->

\linenumbers{}
\modulolinenumbers[3]

\newpage

<!-- Outline

Introduction

1. ECD is an important process, and current empirical and theoretical work suggests it often results in divergence of consumer foraging traits.

2. Ecological consequences of character displacement have received surprisingly little attention. In particular, prior work doesn't allow a proper comparision of the effects of character displacement on community stability.

-> Another paragraph illustrating how the direct and indirect effects of character displacement likely have constrasting effects on stability.

3. Here we fill this knowledge gap. We focus on the ecological side of these dynamics, because the evolutionary aspects have been well characterized, although we show in the appendices that the evolutionary dynamics are as expected (especially for McCann model).

Methods and Results

- Underlying model of consumer-resource dynamics
  1. Explanation of model parameters
  2. Methods for assessing resource abundances and stability.
  3. Trade-off function and definition of character displacement.
  4. Analytical and simulation methods. Briefly mention the different scenarios I'll explore.

- Consumers can forage for resources simultaneously
  1. Functional response equation and illustration of assumptions
  2. Effect on attack rates, resource densities, and stability

- Consumers cannot foraging for resource simultaneously
  1. Functional response equation and illustration of assumptions
  2. Effect on attack rates, resource densities, and stability

- McCann's most realistic functional response
  1. Functional response equation and illustration of assumptions
  2. Effect on attack rates, resource densities, and stability

Discussion

-->

@@

# Abstract

<!--
Ecological character displacement is a key driver of diversification and adaptive radiation; however, there has been little work examining the consequences of character displacement on the dynamics of ecological communities. Given the growing recognition for the role of evolution in structuring ecological communities, understanding how major evolutionary processes impact ecological dynamics is increasingly important. Here, we build a mathematical model to predict how exploitative competition between two consumers for two resources affects the evolution of consumer attack rates, and, in turn, consumer-resource population dynamics. Our model suggested that increasing ecological character displacement decreases the equilibrium density of resources, and with sufficient divergence, results in more oscillatory (destabilizing) consumer-resource dynamics. These results suggest that ecological character displacement can eventually have a negative feedback on food-web persistence in the absence of other processes.
-->

\newpage

# Introduction

<!-- ECD is an important process, and current empirical and theoretical work suggests it often results in divergence of consumer foraging traits.-->

Ecological character displacement is thought to be a key evolutionary process generating biodiversity [@Schluter2000; @Pfennig2010; but see @Stuart2013]. Ecological character displacement describes the “process of phenotypic evolution in a species generated or maintained by [exploitative] resource competition with one or more coexisting species” [@Schluter2000]. Over the past 40 years, a large body of theoretical [e.g. @Lawlor1976; @Abrams1986; @Doebeli1996; @Taper1985] and empirical [reviewed in: @Schluter2000; @Dayan2005; @Stuart2013] work has been generated to understand the scenarios under which exploitative competition for resources leads to the divergence of consumer foraging traits. The general result emerging from these studies is that, if resources are nutritionally substitutable [@Abrams1987; @Fox2008] and there is no other strong source of density dependence acting on consumers [@Abrams1986], then resource competition drives the divergence of consumer foraging traits [@Lawlor1976; @Taper1985). This process is not simply driven by ecological differences, but creates an eco-evolutionary feedback that drives further differentiation. This key insight was made by theoretical models that explicitly included resource dynamics as a mediator of competition in driving evolutionary change [@Lawlor1976; @Abrams1986; @Taper1985].

<!-- Ecological consequences of character displacement have received surprisingly little attention. In particular, prior work doesn't allow a proper comparision of the effects of character displacement on community stability.-->

Although explicit models of resources gave key insights to the evolution of character displacement, the ecological feedback onto consumer-resource dynamics have received surprisingly little attention. This is likely because character displacement has been primarily studied through the lens of competition and coexistence theory [Lawlor1976; Germain2018]. For example, early theoretical work by @Lawlor1976 showed that ecological character displacement promotes coexistence by favoring specialized consumers that experience reduced interspecific competition. On the other hand, specialization often increases a consumer’s ability to capture a specific resource. A well known consequence of increased foraging efficiency is resource suppression, which if sufficient enough, can generate oscillations, and hence less stable, consumer-resource dynamics [Rosenzweig 1971; Murdoch et al. 2003; McCann 2011]. NEED A LAST SENTENCE TO WRAP IT UP, HIGHLIGHTING HOW A FOOD-WEB PERSPECTIVE MAY YIELD NEW ECOLOGICAL INSIGHTS AND GIVE A CONTRASTING PERSPECTIVE TO COEXISTENCE THEORY. PERHAPS A I NEED TO INCLUDE A SENTENCE MID-WAY THROUGH SUGGESTING THAT A FOOD-WEB PERSPECTIVE MAY GIVE QUALITATIVELY DIFFERENT INSIGHTS. IT MAY BE NICE TO BRING IN THE SPATIAL COUPLING WORK OF MCCANN SUGGESTING THAT SPECIALIZATION CAN COUPLE FOOD-WEBS, BUT THIS WAS ONLY WITH ONE TOP PREDATOR (I.E. NO CONSUMERS...LOOK AT OLD TEXT FOR WORDING SUGGESTIONS.)

<!-- older idea...Write 3rd paragraph of intro, highlighting how the direct effects of character displacement, surprisingly, have been ignored. We are currently lacking insight to the relative importance of these direct and indirect effects of character displacement on community dynamics.

<!--
The role of character displacement in governing food-web dynamics remains unclear A food-web perspective, however, has been lacking from theoretical work on the ecological effects of character displacement. , thus the consequences of this process for food-web stabilit

<!--
Despite the key insights made by explicit models of resource dynamics,

gave key insights to the evolutionary dynamics of character displacement, the ecological effects on resources dynamics have received surprisingly little attention. Perhaps this is because the ecological dynamics have been studied primarily through the lens of competition and coexistence thoery

This oversight is unfortunate, given that the stability of consumer-resources is often closely tied with resource abundances.

Although character displacement is the result of an eco-evolutionary feedback, the ecological consequences of this process have received surprisingly little attention [@Germain2018]. Perhaps this is because early theoretical work suggested that character displacement was a potent stabilizing force in ecological communities [@Lawlor1976]. This ecological outcome is intuitive in light of competition theory ---character divergence reduces interspecific competition between consumers, promoting coexistence. On the other hand, specialization often enhances a consumer’s ability to capture a specific resource. A well known consequence of increased foraging efficiency (all else equal) is resource suppression, which if sufficient enough, can generate oscillations, and hence less stable, consumer-resource dynamics (Rosenzweig 1971; Murdoch et al. 2003; McCann 2011).

Moreover, past inferences about the stabilizing effects of character displacement have only been made in communities where both consumers are present. However, I argue that inferences about the effect of character displacement, like the process itself, should only be made when comparing communities in the presence and absence of competition [criteria #3 for character displacement, @Schluter1992; @Schluter2000]. Therefore, the role of ecological character displacement in determining community stability remains unclear.

<!--
However, inferences about the effect of character displacement, like the process itself, can only be made when comparing communities in the presence and absence of competition [criteria #3 for character displacement, @Schluter1992; @Schluter2000]. Current inferences about the stabilizing effects of consumer divergence are only comparisons about communities where there are two consumers, and even though, the focus is only on the stable coexistence of consumer species.

<!--
The consequence of this eco-evolutionary feedback for consumer coexistence has been well characterized (Lawlor and Smith 1976; Germain's work). Ecological character displacement has typically been considered to be a stabilizing factor in communities (Lawlor and Smith 1976). This is because divergence in foraging traits is thought to reduce the strength of interspecific competition between consumers. Reducing the strength of interspecific competition relative to intraspecific competition is a stabilizing force in consumer-resource systems. Consumer specialization on distinct resources suggests that there are niche differences between consumers. Mathematically, this is also often assumed to be the case, because using techniques such as evolutionary invasion analysis describe the ability for a consumer with a mutant allele to increase in abundance when rare, a demographic signature of niche differentiation. Therefore, if both consumers are able to invade when rare, then this suggests that they can coexist.

Character displacement involves evolutionary shifts in character states in response to competition [@Schluter1992; @Schluter2000]. Theoretically, this is determined by comparing character changes in the presence and absence of a competitor. Empirically, this is determined by meeting several criteria that rule out other possible explanations [@Schluter1992; @Schluter2000]. Importantly, current inferences about the effect of character displacement on community stability do not meet this critera. For example, the inference that character displacement enhances species coexistence is only made in a scenario where consumers are competing. This type of stability is important, but it does not enable us to examine the effect of character displacement on community stability. As a consequence, I argue that the effect of character displacement on community stability remains an open question.

Of the prior models of character displacement, only @Lawlor1976 examine the effect of character displacement on the ecological stability of the system. They conducted a detailed theoretical analysis of three different metrics of stability: (i) stable coexistence of consumers, defined as the mutual invasibility of each consumer; (ii) global stability of consumers, defined as the continued coexistence of both consumers when resource dynamics are perturbed; and (iii) local stability, defined as the rate at which the four-species food web returned to equilibrium after a small perturbation to species abundances. Note that the first two metrics correspond to the stable coexistence of consumers, while the third metric measures the stability of the entire food web. @Lawlor1976 found that character displacement enhanced the stable coexistence of consumers, yet had a small negative effect on the local stability of the food web. 
@Lawlor1976's work gave fundamental insight to the effect of character displacement on ecological stability; however, none of their stability analyses compared food webs with and without a competing consumer. Such comparisons are necessary for inferring the effects of ecological character displacement, a point that has been made clear in the criteria to demonstrate character displacement (@Schluter2000; @Schluter1992). In fact, the first two metrics of stability can only be assessed when both consumers are present because it examines the effect of divergence on coexistence, which requires both consumers to be present. In contrast, local stability can be compared in food webs with and without a competing consumer, and thus provides a useful benchmark for examining the effects of character displacement on food-web stability.

<!-- -> Another paragraph illustrating how the direct and indirect effects of character displacement likely have constrasting effects on stability. -->

<!--
Indeed, ecological character displacement often results in the evolution of characters that enhance a consumer’s ability to capture resources, which tends to destabilize consumer-resource interactions (Murdoch et al. 2003; McCann 2011). This occurs because increases in a consumer’s attack rate results in the resource being suppressed well below its carrying capacity, which if sufficient enough, can generate oscillations, and hence less stable, consumer-resource dynamics (Rosenzweig 1971; Murdoch et al. 2003). Experimental decreases in interaction strength between consumers and resources has been shown to promote coexistence between consumers and resources (Luckinbill 1973). Moreover, there is broad support that specialized consumers, since they are tightly coupled to their resources, tend to exhibit consumer-resource cycles, whereas this is not the case for generalists (Murdoch et al. 2002). Therefore, although evolutionary responses to exploitative competition may reduce the strength of interactions between consumers, they ultimately increase the strength of interactions with their resources which can destabilize the direct consumer-resource interaction (rather than the indirect consumer-consumer interaction). While prior models of character displacement have included more realistic functional responses and spatial relationships between consumers and resources, they have focused more on whether these factors alter patterns of character displacement rather than the effects of character displacement on the stability of the ecological community. As a consequence, it is currently unclear whether ecological character displacement has a net stabilizing or destabilizing effect on ecological communities.

<!-- Here we fill this knowledge gap. We focus on the ecological side of these dynamics, because the evolutionary aspects have been well characterized, although we show in the appendices that the evolutionary dynamics are as expected (especially for McCann model). -->

Here, I address this knowledge gap by building a mathematical model to examine how ecological character displacement affects consumer-resource dynamics in a food-web context. I address two questions: (1) How does ecological character displacement affect resource abundances? (2) How does character displacement affect food-web stability? To test the generality of these ecological effects, I examined three different foraging scenarios: (i) resources occur in the same habitat and consumers exhibit a linear functional response (e.g. @MacArthur1972; @Lawlor1976); (ii) resources occur in different habitats and consumers exhibit a linear functional response (e.g. @Lawlor1976); (iii) resources occur in distinct habitats and consumers exhibit a more realistic functional response [@McCann2005]. I also explore the contingency of these effects to different assumptions about evolutionary trade-offs in consumer foraging traits. I focus on the ecological side of these dynamics because the evolutionary aspects have been well characterized [@Lawlor1976; @Taper1985; @Abrams1986]. I show that the consequences of character displacement on food-web dynamics are not always intuitive and depend on the foraging scenario and tradeoff in consumer traits.

# Methods and Results

## Underlying Food-web Dynamics

<!-- Explanation of model parameters -->

To examine how ecological character displacement affects consumer-resource interactions and food-web dynamics, I analyzed a continuous-time model of two consumers ($C_1,C_2$) competing for two shared resources ($R_1,R_2$):

$$\frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{21}(R_1)C_2,$$
$$\frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-F_{12}(R_2)C_1-F_{22}(R_2)C_2,$$
$$\frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{12}F_{12}(R_2)C_1-m_1C_1,$$
$$\frac{dC_2}{dt}=e_{21}F_{21}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2,$$

where $r_i$ represents the intrinsic growth rate of resource $i$, $K_i$ represents the carrying capacity of resource $i$, $e_{ji}$ represents the conversion efficiency of resource $i$ into consumer $j$, and $m_j$ represents the mortality rate of consumer $j$. $F_{ji}(R_i)$ represents consumer $j$'s feeding rate (i.e. functional response) on resource $i$. This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size, see @Taper1985 for an example). In the scenario where consumer feeding rates increase linearly with resource density, such that $F_{ji}(R_i)=a_{ji}R_i$ where $a_{ji}$ is the attack rate of consumer $j$ on resource $i$, then this model becomes MacArthur's model of resource competition (@MacArthur1972). The evolution of consumer attack rates in this model (and several extensions) have been analyzed in detail (@Lawlor1976; @Abrams1986), with the general result being divergent character displacement. I say consumers have undergone divergent character displacement if there evolved attack rates are more specialized, defined as $|\frac{a_{ii}}{a_{ii}+a_{ij}}-0.5|$, when evolving with a competing consumer. I focus here on the ecological consequences of this adaptive divergence.

<!-- Methods for assessing resource abundances and stability. -->

To investigate the ecological consequences, I studied the effect of adaptive divergence in consumers on resource densities and the stability of the entire food web at equilibrium. An equilibrium is reached when the four rates of change in the above equations are 0, and solving the system at this point gives equilibrium densities for each resource ($\hat R_1,\hat R_2$) and consumer ($\hat C_1,\hat C_2$). At the coexistence equilibrium (i.e., where $\hat R_1,\hat R_2, \hat C_1,\hat C_2 > 0$), I calculated resource densities and performed a linear stability analysis [@Otto2007]. This stability analysis derives the dominant (largest in absolute value) eigenvalue, $\lambda$, which determines whether (and how readily) the food web, when perturbed a small amount, will return to equilibrium (see supplementary material for derivation). Together, these two measures indicate how ecological character displacement affects resource densities and food-web stability.

<!-- Explain analytical methods -->

I examine the effects of this inevitable divergence in attack rates on food-web dynamics by (i) deriving analytical expressions for the relationship between attack rates and food-web dynamics and (ii) simulating the effects of competition on the eco-evolutionary dynamics of consumers and resources. To gain analytical insight, I assume that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $m=m_1=m_2$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). While this scenario is arguably simplistic, it allows me to isolate the effects of ecological character displacement. <!--, rather than any dependence on inherent asymmetries between consumers or resources.-->

<!-- Explain simulations -->

To simulate eco-evolutionary dynamics, I used an Adaptive Dynamics approach. At each evolutionary time step, I created a mutant consumer by randomly choosing a consumer and modifying its attack rate on one resource by either subtracting or adding a small constant (0.001 in the following simulations) with equal probability. The mutant's attack rate on the other resource was determined by a tradeoff, such that $\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff [@Sargent2006]. This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher, on average, than the extremes (when $n>1$, green line in Fig. \ref{fig:Tradeoff}) or, conversely, where the two extremes are higher, on average, than intermediate investments (when $n<1$, orange line Fig. \ref{fig:Tradeoff}). When $n=1$, the tradeoff function is linear, and all combinations of $a_{ii}$ and $a_{ij}$ have the same total attack rate (blue line in Fig. \ref{fig:Tradeoff}). Assuming the mutant consumer was rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource dynamics to reach a steady state. I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy [ESS, @Smith1973] or an evolutionary limit (e.g. $\frac{a_{ii}}{a_{ii}+a_{ij}}$ is constrained to a maximum of 1 and minimum of 0).

<!-- Redundant, removing for now
In what follows, I examine the ecological consequences of character displacement under three different attack-rate tradeoffs (n>1, n=1, n<1) and in three different foraging scenarios: (i) resources occur in the same habitat and consumers exhibit a linear functional response (e.g. @MacArthur1972); (ii) resources occur in different habitats and consumers exhibit a linear functional response (e.g. @Lawlor1976); (iii) resources occur in distinct habitats and consumers exhibit a more realistic functional response [@McCann2005]. I show that the consequences of character displacement for food-web dynamics are not intuitive and depend critically on the foraging scenario and tradeoff in consumer attack rates. 
-->

```{r Tradeoff, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:Tradeoff}Tradeoff forms for consumer attack rates."}
tradeoff_plot <- bind_rows(concaveDN_df, linear_df, concaveUP_df) %>%
  ggplot(., aes(x=aii, y=aij, color = factor(n))) + 
  geom_line() + 
  xlab(expression(a[ii])) +
  ylab(expression(a[ij])) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)

@@

## Consumers can forage for both resources simultaneously

<!-- Functional response equation and illustration of assumptions already given in previous section -->

<!-- Redundant, removing for now
The evolution of consumer attack rates in this model has been analyzed in detail [@Lawlor1976; @Abrams1986], with the general result being divergent character displacement. -->
When both consumers and resources are present, the density of resources at equilibrium are equivalent ($\hat R = \hat R_1 = \hat R_2$) and are determined by the following equation (derivation in supplementary material):

$$\hat{R}=\frac{1}{a_{ii}+a_{ij}}\cdot\frac{m}{e}$$
A key determinant of resource density in this model is the consumer's total attack rate, $a_{ii}+a_{ij}$. Therefore, the ecological consequences of character displacement depend on how the tradeoff influences the evolution of consumer attack rates.

My simulations show that the shape of the attack-rate tradeoff qualitatively affects the relationship between character displacement and resource density (Fig. \ref{fig:MacArthur_Resources}). For example, if consumer's are constrained by a linear tradeoff (blue lines), then there is no net change in total attack rate (Fig. \ref{fig:MacArthur_Resources}A) and character displacement has no effect on resource densities (Fig. \ref{fig:MacArthur_Resources}B). If the tradeoff is concave down (green lines), then resource abundances can actually increase under character displacement (Fig. \ref{fig:MacArthur_Resources}B). This is because the total attack rate of consumers is maximized at intermediate values ($a_{ii}=a_{ij}$) and decreases as consumers diverge (Fig. \ref{fig:MacArthur_Resources}A). When the tradeoff is concave up (orange lines), character displacement suppresses resource densities due to the increase in total attack rates (Fig. \ref{fig:MacArthur_Resources}A,B). Although the equation I derived for resource density was for the scenario where both consumers and both resources are present, it accurately predicts the density of resources when a single consumer reaches its ESS (triangles on respective colored lines in Fig. \ref{fig:MacArthur_Resources}B). This is because a single consumer evolves to be a generalist that has equal attack rates on each resource (triangles at 0.5 along x-axis in Fig. \ref{fig:MacArthur_Resources}A), resulting in equivalent resource densities.
